\papertitle

### 2017-03-27 article-icmc2017.tex
save work

The main problem that arises is that composers may specify relationships between processes, but such processes may not always be active. 
We start by making the case for an environment associated with the dataflow graph that contains the values of the inputs and outputs.
Then, since a graph may not always be fully active, we show that it is also meaningful to have implicit connections between nodes of the graph, that will use the environment.
This allows dynamic routing of the processes according to their temporal order during execution, and extends routing to pattern-matched elements of the environment instead of single values.
We enumerate different behaviours regarding the input and output parameters when the nodes are implicitely connected at runtime using the environment: strict, glutton, parallelized, and replaced. 
Additionnaly, when the node are explicitely connected, we introduce a delay mechanism that allows a computation to happen on data previously produced by another node of the graph, instead of reading the current tokens.

@@

Dynamicity in dataflows is generally exprimed on two independent aspects : dynamicity of the data, and of the topology.
The first relates to the variability on the streams of tokens, while the second is about changes to the structure of the graph. 
Boolean parametric dataflows\cite{bempelis2015boolean} have been proposed to solve dynamicity of topology, by introducing conditionals at the edges.

\subsection{Temporal formalism}
This work implicitely uses the OSSIA formalism~\cite{celerier2015ossia} to provide primitives relative to the evolution of time: time constraints (horizontal lines), time nodes (vertical lines). 
The elements of its syntax are presented in fig.~\ref{fig.iscore-example}.

@@

This work extends it with a permanent external environment, that contains a mapping of key-value pairs.
Keys are specified with OSC-like\cite{Freed09featuresand} adresses: \lstinline|/foo/bar| and values can be any kind of data: numbers, strings, etc. 
The values of the environment may vary independently of the execution of the program: for instance elements of the environment may be changed asynchronously through the network, graphical widgets or physical controls.

We then have to provide a mean to use the values of this environment to the dataflow.
Thus, we extend the inputs and outputs ports with the following expliciteness semantic:

\begin{itemize}
  \item An explicit port has been connected manually to another port.
        An input port may only read tokens from the output port it is connected to.
  \item An implicit port has not been connected to another port. 
        It may read and write to any number of adresses from the environment to perform its work.
        Every write done by a node is inserted in a sub-environment scoped to the duration of the tick.
        It is up to the developer of the music software to choose whether the requested addresses for each node are specified statically, that is, in the definition of the dataflow, or dynamically.
        Dynamic behaviour would allow to receive input from a random output at each tick for instance, but makes the program harder to comprehend for authors.
\end{itemize}

During a tick, adresses have access to two sets of values, corresponding to scopes: 
\begin{itemize}
  \item Global scope: The values that were in the environment at the beginning of the tick.
  \item Local scope: The values produced by previous implicit output ports in the dataflow.
        Section~\ref{sec.relationships} presents the various policies that nodes can leverage to 
        use these values. 
        These values are committed to the environment at the end of a tick.
\end{itemize}

@@

It is up to the implementor of nodes of the graph to choose whether a port can work: 
\begin{itemize}
  \item When connected explicitely,
  \item With the environment produced from previous nodes,
  \item With the global environment,
  \item With both of the previous.
\end{itemize}

This mechanism allows for new behaviours during the creation of a dataflow program.
For instance, instead of accepting data from a single source, a port can specify an input pattern such as: \lstinline|/foo.*/bar/b[a-zA-Z]|, as proposed in the OSC specification.

Upon execution, the input stream of the port would contain the list of the matching adresse

@@

% TODO sample-precision of the activation / deactivation of a process / node
% TODO 
\section{Relationships}
\label{sec.relationships}
We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens, whether the production is specified implicitely or explicitely.

@@

We will get the following behaviours during each slice: 
\begin{itemize}
  \item During $t_0$: $\emptyset$.
  \item During $t_1$: $f_2 \circ f_1 $ or $f_1 \circ f_2$ depending on the node ordering.
  \item During $t_2$: $\emptyset$. 
\end{itemize}

The strict execution may happen explicitely or implicitely.

@@

If we take the same case than previously, the behaviour is:
\begin{itemize}
  \item $t_0$: $f_1$.
  \item $t_1$: $f_2 \circ f_1$ or $f_1 \circ f_2$.
  \item $t_2$: $f_2$. 
\end{itemize}

@@

The delayed execution has be specified explicitely for the tokens to be bufferised.

If we have a delay connection from $f_1$ to $f_2$, the first call to $f_2$ will use the first value that was produced by $f_1$.
If the execution order happens to be reversed, such as in the case of fig.~\ref{fig.simple-reversed}, nothing will be read, and the execution will only happen during $t_1$.
\subsection{Parallelization and mixing}
The next nodes are run with the same input environment, and summed.

@@

\begin{itemize}
  \item $t_0$: $f_1$.
  \item $t_1$: $s(f_1, f_2)$ where $s$ is the summing operation, and $f_1$, $f_2$ are ran with the same input environment
  \item $t_2$: $f_2$. 
\end{itemize}

The parallelized execution may happen explicitely or implicitely.

@@

The replaced execution may happen explicitely or implicitely.

@@

Finally, default behaviours can be encoded in any hierarchical unit such as the global scope or each time flow could be set-up with a default connection type.

We propose a default glutton behaviour; however usability studies should be led to determine the most user-friendly default behaviour.

\section{Default behaviours for orders}
\label{sec.order}
We assess here the various mechanisms possible to specify the dependencies between nodes of the dataflow.
Information provided provided in the timeflow can be leveraged to connect the nodes at runtime in a fashion 
that would make sense to composers \& authors.

@@

Connections between ports imply an ordering between nodes, hence we only consider the ordering between implicitely connected nodes of the dataflow.

\subsection{Manual ordering}
In this case, the author explicitely specifies an order by setting up an edge between two nodes. 
This is the slowest process, but which gives the most precise specification.

@@

%Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
%\section{Implementation}
\section{Conclusion}
Throughought this paper, the effects of adding macro-temporal semantics to dataflows were studied. 
The first step is to introduce an environment that allows execution of a program to continue if not all the nodes of a dataflow are currently being executed.
Then, we study the new possibilites offered by the presence of the environment by the different ways to interpret the absence of execution of a node.
This includes matching nodes together through pattern matching, and distinct policies of execution that can take into account the evolution of the program in time.

An implementation is being developed as an extension to the i-score software. 
This implementation uses PureData through libpd to provide function execution with multiple ports
with primitives useful for musical creation.

%Additional consideration: see PLOP 2006 article (multi-stream, etc), sample accuracy, etc

>>>>

The main problem that arises is that composers may specify relationships between processes, but such processes may not always be active. 
We start by making the case for an environment associated with the dataflow graph that contains the values of the inputs and outputs.
Then, since a graph may not always be fully active, we show that it is also meaningful to have implicit connections between nodes of the graph, that will use the environment.
This allows dynamic routing of the processes according to their temporal order during execution, and extends routing to pattern-matched elements of the environment instead of single values.
We enumerate different behaviors regarding the input and output parameters when the nodes are implicitly connected at runtime using the environment: strict, glutton, parallelized, and replaced. 
Additionally, when the node are explicitly connected, we introduce a delay mechanism that allows a computation to happen on data previously produced by another node of the graph, instead of reading the current tokens.

@@

Dynamicity in dataflows is generally separated in two independent aspects : dynamicity of the data, and of the topology.
The first relates to the variability on the streams of tokens, while the second is about changes to the structure of the graph. 
Boolean parametric dataflows\cite{bempelis2015boolean} have been proposed to solve dynamicity of topology, by introducing conditionals at the edges.

\subsection{Temporal formalism}
This work implicitly uses the OSSIA formalism~\cite{celerier2015ossia} to provide primitives relative to the evolution of time: time constraints (horizontal lines), time nodes (vertical lines). 
The elements of its syntax are presented in fig.~\ref{fig.iscore-example}.

@@

This work extends it with a permanent external environment, that contains a mapping of key-value pairs.
Keys are specified with OSC-like\cite{Freed09featuresand} addresses: \lstinline|/foo/bar| and values can be any kind of data: numbers, strings, etc. 
The values of the environment may vary independently of the execution of the program: for instance elements of the environment may be changed asynchronously through the network, graphical widgets or physical controls.

We then have to provide a mean to use the values of this environment to the dataflow.
Thus, we extend the inputs and outputs ports with the following explicitness semantic:

\begin{itemize}
  \item An explicit port has been connected manually to another port.
        An input port may only read tokens from the output port it is connected to.
  \item An implicit port has not been connected to another port. 
        It may read and write to any number of addresses from the environment to perform its work.
        Every write done by a node is inserted in a sub-environment scoped to the duration of the tick.
        It is up to the developer of the music software to choose whether the requested addresses for each node are specified statically, that is, in the definition of the dataflow, or dynamically.
        Dynamic behavior would allow to receive input from a random output at each tick for instance, but makes the program harder to comprehend for authors.
\end{itemize}

During a tick, addresses have access to two sets of values, corresponding to scopes: 
\begin{itemize}
  \item Global scope: The values that were in the environment at the beginning of the tick.
  \item Local scope: The values produced by previous implicit output ports in the dataflow.
        Section~\ref{sec.relationships} presents the various policies that nodes can leverage to 
        use these values. 
        These values are committed to the environment at the end of a tick.
\end{itemize}

@@

It is up to the implementor of nodes of the graph to choose whether a port can work: 
\begin{itemize}
  \item When connected explicitly,
  \item With the environment produced from previous nodes,
  \item With the global environment,
  \item With both of the previous.
\end{itemize}

This mechanism allows for new behaviors during the creation of a dataflow program.
For instance, instead of accepting data from a single source, a port can specify an input pattern such as: \lstinline|/foo.*/bar/b[a-zA-Z]|, as proposed in the OSC specification.

Upon execution, the input stream of the port would contain the list of the matching addresses.

@@

% TODO sample-precision of the activation / deactivation of a process / node
% TODO 
\section{Relationships}
\label{sec.relationships}
We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens, whether the production is specified implicitly or explicitly

@@

We will get the following behaviors during each slice: 
\begin{itemize}
  \item During $t_0$: $\emptyset$.
  \item During $t_1$: $f_2 \circ f_1 $ or $f_1 \circ f_2$ depending on the node ordering.
  \item During $t_2$: $\emptyset$. 
\end{itemize}

The strict execution may happen explicitly or implicitly.

@@

If we take the same case than previously, the behavior is:
\begin{itemize}
  \item $t_0$: $f_1$.
  \item $t_1$: $f_2 \circ f_1$ or $f_1 \circ f_2$.
  \item $t_2$: $f_2$. 
\end{itemize}

@@

The delayed execution has be specified explicitly for the tokens to be bufferised.

If we have a delay connection from $f_1$ to $f_2$, the first call to $f_2$ will use the first value that was produced by $f_1$.
If the execution order happens to be reversed, such as in the case of fig.~\ref{fig.simple-reversed}, nothing will be read, and the execution will only happen during $t_1$.
\subsection{Parallelization and reduction}
The next nodes are run with the same input environment, and summed:

\begin{itemize}
	\item $t_0$: $f_1$.
	\item $t_1$: $s(f_1, f_2)$ where $s$ is the reduction operation, and $f_1$, $f_2$ are ran with the same input environment.
	\item $t_2$: $f_2$. 
\end{itemize}

@@

The parallelized execution may happen explicitly or implicitly

@@

The replaced execution may happen explicitly or implicitly.

@@

Finally, default behaviors can be encoded in any hierarchical unit such as the global scope or each time flow could be set-up with a default connection type.

We propose a default glutton behavior; however usability studies should be led to determine the most user-friendly default behavior.

\section{Default behaviors for orders}
\label{sec.order}
We assess here the various mechanisms possible to specify the dependencies between nodes of the dataflow.
Information provided provided in the timeflow can be leveraged to connect the nodes at runtime in a fashion 
that would make sense to composers \& authors.

@@

Connections between ports imply an ordering between nodes, hence we only consider the ordering between implicitly connected nodes of the dataflow.

\subsection{Manual ordering}
In this case, the author explicitly specifies an order by setting up an edge between two nodes. 
This is the slowest process, but which gives the most precise specification.

@@

%Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
%\section{Implementation}
\section{Conclusion}
Throughout this paper, the effects of adding macro-temporal semantics to dataflows were studied. 
The first step is to introduce an environment that allows execution of a program to continue if not all the nodes of a dataflow are currently being executed.
Then, we study the new possibilities offered by the presence of the environment by the different ways to interpret the absence of execution of a node.
This includes matching nodes together through pattern matching, and distinct policies of execution that can take into account the evolution of the program in time.

An implementation is being developed as an extension to the i-score software. 
This implementation uses PureData via libpd to provide function execution with multiple ports
with primitives useful for musical creation.

A further goal for this work is to extend the definition not only to work with a single variable of time, but to any kind of activation mechanism: for instance, evolution of a position in a 3D space, etc.
%Additional consideration: see PLOP 2006 article (multi-stream, etc), sample accuracy, etc
